### 2021-01-08 paper/sex_selection_spacing-ver5-demography.tex
Version 5 submitted to Demography

\title{Birth Spacing and Fertility in the Presence of Son Preference and Sex-Selective Abortions:
India's Experience Over Four Decades%
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to Monica Das Gupta, Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, three anonymous reviewers, and seminar participants at University of 
Copenhagen, University of Michigan, University of Washington, University of Aarhus, the 
Fourth Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful 
suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support for development of the method from the University of Washington Royalty 
Research Fund and the Development Research Group of the World Bank is gratefully 
acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
}
}

\author{Claus C P\"ortner\\
    Department of Economics\\
    Albers School of Business and Economics\\
    Seattle University, P.O. Box 222000\\
    Seattle, WA 98122\\
    \href{mailto:cportner@seattleu.edu}{\texttt{cportner@seattleu.edu}}\\
    \href{http://www.clausportner.com}{\texttt{www.clausportner.com}}\\
    \& \\
    Center for Studies in Demography and Ecology \\
    University of Washington\\ \vspace{2cm}
    }

@@

\setcounter{page}{-1}
\maketitle
\thispagestyle{empty}

% \setcounter{page}{0}

\newpage
\thispagestyle{empty}
\doublespacing

\begin{abstract}

% Demography abstract
\noindent 
% In India, the most substantial lengthening in Hindu women's birth intervals over the past 
% four decades has occurred for the women most likely to use sex-selective abortions. 
% For well-educated women with two girls, the median birth interval before the third child 
% has increased by nearly 15 months---and the 75th percentile interval by 21 months---as the 
% sex ratio has approached 80\% boys. 
% Because of the rapid lengthening in spacing, standard fertility rates have significantly 
% overestimated how fast cohort fertility fell. 
% Despite convergence, cohort fertility is still 10\%--20\% higher than the 
% fertility rate and above replacement level for all but the best-educated urban women. 
% With the lengthening from sex selection, some women show a reversal of the traditional 
% spacing pattern; instead of the shortest birth interval in the absence of sons, they now 
% have the longest. 
% Finally, infant mortality has declined over time for all groups, but the decline was 
% fastest for the less educated. 
% Short birth spacing is still associated with higher mortality, although considerably less 
% so for the best-educated women. 
% There is no evidence that the increasing use of sex selection is associated with higher 
% infant mortality. 
% Neither does it appear that the use of sex selection is declining.

% \noindent Over the past four decades, the Hindu women in India most likely to use 
% sex-selective abortions---well-educated women with no sons---had the most substantial 
% lengthening of birth intervals and the most biased sex ratios. 
% As a result, we now see cases that reverse the traditional spacing pattern, with some
% women with no sons having longer birth intervals than those with sons. 
% Those least likely to use sex-selective abortions---less-educated women in rural 
% areas---still follow the traditional pattern of short spacing when they have girls, with 
% only limited evidence of sex selection. 
% Because of the rapid lengthening in spacing, the standard fertility rates substantially 
% overestimated how fast cohort fertility fell. 
% Despite a convergence, cohort fertility is still 10\%--20\% higher than the fertility rate 
% and above replacement level for all but the best-educated urban women. 
% Infant mortality has declined substantially over time for all groups, with the fastest 
% decline among the less educated. 
% Short birth spacing is still associated with higher mortality, although considerably less 
% so for the best-educated women. 
% There is no evidence that repeated sex-selective abortions are associated with higher 
% infant mortality for the child eventually born. 
% Finally, it does not appear that the use of sex selection is declining.

% \noindent India has experienced an increasingly male-biased sex ratio at birth following
% the introduction of prenatal sex determination techniques. 
% I examine three questions.
% First, how did birth spacing change with this and other secular developments? 
% Second, did changes in birth spacing make us overestimate the decline in cohort fertility? 
% Finally, did the advent of sex-selective abortions change the relationship between birth 
% spacing and infant mortality? 
% Well-educated Hindu women with no sons had the most substantial lengthening of birth 
% intervals and the most male-biased sex ratios, likely from sex-selective abortions. 
% Consequently, some women with no sons now have longer birth intervals than those with sons, 
% reversing India's traditional spacing pattern.
% Less-educated women continued short spacing when they have girls, with only limited
% evidence of male-biased sex ratios. 
% Because of the overall lengthening in spacing, period fertility rates substantially 
% overestimated how fast cohort fertility fell. 
% Furthermore, predicted cohort fertility remains above replacement for all but the 
% best-educated urban women. 
% If lengthening of birth intervals occurs from repeated abortions, the associated
% short pregnancy spacing may counteract any positive effects of longer birth spacing. 
% There is no evidence of this effect on infant mortality. 
% Finally, judging from sex ratios, sex-selective abortion use is not declining.

\noindent Following the arrival of prenatal sex-determination technologies in the mid-1980s, 
India has experienced an increasingly male-biased sex ratio at birth, presumably from sex-selective 
abortions. 
Abortions lengthen birth intervals, but we know little about how birth spacing has changed or the 
effects of these changes. 
This paper shows that, although the length of birth intervals increased overall, well-educated women 
with no sons had the most substantial lengthening---and the most male-biased sex ratios. 
Furthermore, most of these changes took place immediately after the introduction of prenatal 
sex-determination technologies. 
Consequently, some women without sons now have longer birth intervals than those with sons, reversing 
India's traditional spacing pattern. 
Less-educated women continue short birth spacing when they have no sons, with only limited evidence 
of male-biased sex ratios. 
Because of the rapid lengthening of birth intervals, period fertility rates substantially overestimated 
how fast cohort fertility fell. 
Moreover, predicted cohort fertility is still 10\%--20\% above the period fertility rate. 
If lengthening of birth intervals arise from repeated abortions, the associated short pregnancy spacing 
may counteract any positive effects of longer birth spacing. 
There is, however, no evidence of this effect on infant mortality. 
Finally, judging from sex ratios, sex-selective abortion use is not declining.

>>>>

\title{Birth Spacing and Fertility in the Presence of Son Preference and Sex-Selective Abortions:
India's Experience Over Four Decades%
% \protect\thanks{%
% I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
% I owe thanks to Monica Das Gupta, Shelly Lundberg, Daniel Rees, David Ribar, 
% Hendrik Wolff, three anonymous reviewers, and seminar participants at University of 
% Copenhagen, University of Michigan, University of Washington, University of Aarhus, the 
% Fourth Annual Conference on Population, Reproductive Health, 
% and Economic Development, and the Economic Demography Workshop for helpful 
% suggestions and comments.
% I would also like to thank Nalina Varanasi for research assistance.
% Support for development of the method from the University of Washington Royalty 
% Research Fund and the Development Research Group of the World Bank is gratefully 
% acknowledged.
% The views and findings expressed here are those of the author and
% should not be attributed to the World Bank or any of its member countries.
% Partial support for this research came from a Eunice Kennedy Shriver National
% Institute of Child Health and Human Development research infrastructure grant,
% 5R24HD042828, to the Center for Studies in Demography and Ecology at the
% University of Washington.
% }
}

\author{}

% \author{Claus C P\"ortner\\
%     Department of Economics\\
%     Albers School of Business and Economics\\
%     Seattle University, P.O. Box 222000\\
%     Seattle, WA 98122\\
%     \href{mailto:cportner@seattleu.edu}{\texttt{cportner@seattleu.edu}}\\
%     \href{http://www.clausportner.com}{\texttt{www.clausportner.com}}\\
%     \& \\
%     Center for Studies in Demography and Ecology \\
%     University of Washington\\ \vspace{2cm}
%     }

@@

% \setcounter{page}{-1}
\maketitle
\thispagestyle{empty}

\setcounter{page}{0}

% \newpage
\thispagestyle{empty}
\doublespacing

\begin{abstract}
\noindent Following the arrival of prenatal sex-determination technologies in the mid-1980s, 
India has experienced an increasingly male-biased sex ratio at birth, presumably from sex-selective 
abortions. 
Abortions lengthen birth intervals, but we know little about how birth spacing has changed or the 
effects of these changes. 
This paper shows that, although the length of birth intervals increased overall, well-educated women 
with no sons had the most substantial lengthening---and the most male-biased sex ratios. 
Furthermore, most of these changes took place immediately after the introduction of prenatal 
sex-determination technologies. 
Consequently, some women without sons now have longer birth intervals than those with sons, reversing 
India's traditional spacing pattern. 
Less-educated women continue short birth spacing when they have no sons, with only limited evidence 
of male-biased sex ratios. 
Because of the rapid lengthening of birth intervals, period fertility rates substantially overestimated 
how fast cohort fertility fell. 
Moreover, predicted cohort fertility is still 10\%--20\% above the period fertility rate. 
If lengthening of birth intervals arise from repeated abortions, the associated short pregnancy spacing 
may counteract any positive effects of longer birth spacing. 
There is, however, no evidence of this effect on infant mortality. 
Finally, judging from sex ratios, sex-selective abortion use is not declining.
